SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-05-09 paper/paper.tex
regression: add generic QC shrink instances.

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for
automatic test-case generation.  Because QuickCheck performs random testing,
some of the counterexamples discovered are very large.  QuickCheck provides an
interface for the user to write shrink functions to attempt to reduce the size
of counterexamples.  Hand-written implementations of shrink can be complex,
inefficient, and consist of significant boilerplate code.  Furthermore,
shrinking is only one aspect in debugging: counterexample generalization is the
process of extrapolating from individual counterexamples to a class of
counterexamples, often requiring a flash of insight from the programmer.  To
improve counterexample reduction and generalization, we introduce
\emph{SmartCheck}.  SmartCheck is a debugging tool that reduces algebraic data
using generic search heuristics to efficiently find smaller counterexamples.  In
addition to shrinking, SmartCheck also automatically generalizes counterexamples
to formulas representing classes of counterexamples.  SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

%
\begin{samepage}
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
\end{samepage}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the property, so that the user can focus on the heart of the
problem in a class of counterexamples.

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle and generically minimizes and generalizes the counterexample.

@@

After presenting some preliminary definitions in
Section~\ref{sec:preliminaries}, in Section~\ref{sec:shrinking}, we describe the
generic shrinking algorithm implemented.  The algorithm is particularly novel in
that it generates new random sub-values during the reduction phase (see related
work in Section~\ref{sec:related}).

@@

toList :: T -> [[Int16]]
toList (T i0 i1 i2 i3 i4) =
  (map . map) fromIntegral [i0, i1, i2, i3, i4]

pre :: T -> Bool
pre t = all ((>) 256 . sum) (toList t)

@@

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing five lists of
signed 8-bit integers.  %% For our purposes, it is not important what \ttp{T}
%% models, but for concreteness, suppose it models the values in five linked lists.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable on first glance, there are two
problems with it.  First, the result of (\ttp{shrink t}) is null if any list
contained in \ttp{t} is null, in which case \ttp{t} will not be shrunk.  More
troublesome is that with QuickCheck's default \ttp{Arbitrary} instances, the
length of potential counterexamples returned by \ttp{shrink} can easily exceed
$10^{10}$, which is an intractable number of tests for QuickCheck to analyze.
The reason for the blowup is that shrinking a list \ttp{[a]} produces a list
\ttp{[[a]]}, a list of lists; each list is a new shrinking of the original list.
Then, we take the cross-product of the generated lists.  For the example above,
with some counterexamples, the shrinking stage may appear to execute for hours,
consuming ever more memory, without returning a result.

The first way we will control the complexity is focusing on structurally
shrinking the data type rather than shrinking the base types.  We prevent
QuickCheck from using its default shrink instance for \ttp{Int16} by defining a
newtype:
%
\begin{code}
newtype J = J { getInt :: Int16 }
\end{code}
%
\noindent
Next, a programmer might try to control the complexity by truncating lists using
truncation where
%
\begin{code}
take n ls
\end{code}
%
returns the first $n$ elements \ttp{ls}.  The trade-off is quicker shrinking
with a lower probability of finding a smaller counterexample.  For example, we
might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space, but the downside is that potentially smaller counterexamples may be
omitted.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int}
  or \ttp{Char}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, the anonymous
reviewers of ICFP'13 (from which an early draft of this paper was rejected), and
especially John Hughes for reading an earlier draft of this paper.  Their deep
insights and bug catches dramatically improved both the form and content.

>>>>

\begin{abstract}
QuickCheck is a powerful library for automatic test-case generation.  Because
QuickCheck performs random testing, some of the counterexamples discovered are
very large.  QuickCheck provides an interface for the user to write shrink
functions to attempt to reduce the size of counterexamples.  Hand-written
implementations of shrink can be complex, inefficient, and consist of
significant boilerplate code.  Furthermore, shrinking is only one aspect in
debugging: counterexample generalization is the process of extrapolating from
individual counterexamples to a class of counterexamples, often requiring a
flash of insight from the programmer.  To improve counterexample reduction and
generalization, we introduce \emph{SmartCheck}.  SmartCheck is a debugging tool
that reduces algebraic data using generic search heuristics to efficiently find
smaller counterexamples.  In addition to shrinking, SmartCheck also
automatically generalizes counterexamples to formulas representing classes of
counterexamples.  SmartCheck has been implemented for Haskell and is freely
available.
\end{abstract}

@@

%
\begin{samepage}
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
\end{samepage}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the property, so that the user can focus on the heart of the
problem in a class of counterexamples.  Given a program and large
counterexample, \emph{SmartCheck} returns such a formula.

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed SmartCheck.  SmartCheck takes a counterexample produced by some
oracle and generically minimizes and generalizes the counterexample.

@@

After presenting some preliminary definitions in
Section~\ref{sec:preliminaries}, in Section~\ref{sec:shrinking}, we describe
SmartCheck's generic counterexample reduction algorithm.  The algorithm is
particularly novel in that it generates new random sub-values during the
reduction phase (see related work in Section~\ref{sec:related}).

@@

toList :: T -> [[Int16]]
toList (T i0 i1 i2 i3 i4) = [i0, i1, i2, i3, i4]

pre :: T -> Bool
pre t = all ((< 256) . sum) (toList t)

@@

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing five lists of
signed 16-bit integers.  %% For our purposes, it is not important what \ttp{T}
%% models, but for concreteness, suppose it models the values in five linked lists.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T i0 i1 i2 i3 i4) =
  [ T a b c d e | a <- shrink i0, b <- shrink i1
                , c <- shrink i2, d <- shrink i3
                , e <- shrink i4 ]
\end{code}
%
\noindent
While the above definition appears reasonable on first glance, there are two
problems with it.  First, the result of (\ttp{shrink t}) is null if any list
contained in \ttp{t} is null, in which case \ttp{t} will not be shrunk.  More
troublesome is that with QuickCheck's default \ttp{Arbitrary} instances, the
length of potential counterexamples returned by \ttp{shrink} can easily exceed
$10^{10}$, which is an intractable number of tests for QuickCheck to analyze.
The reason for the blowup is that shrinking a list \ttp{[a]} produces a list
\ttp{[[a]]}, a list of lists; each list is a new shrinking of the original list.
Then, we take the cross-product of the generated lists.  For the example above,
with some counterexamples, the shrinking stage may appear to execute for hours,
consuming ever more memory, without returning a result.

The first way we will control the complexity is focusing on structurally
shrinking the data type rather than shrinking the base types.  We prevent
QuickCheck from using its default shrink instance for \ttp{Int16} by defining a
newtype:
%
\begin{code}
newtype J = J \{ getInt :: Int16 \}
\end{code}
%
\noindent
Next, a programmer might try to control the complexity by truncating lists using
truncation where
%
\begin{code}
take n ls
\end{code}
%
returns the first $n$ elements \ttp{ls}.  The trade-off is quicker shrinking
with a lower probability of finding a smaller counterexample.  For example, we
might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space, but the downside is that potentially smaller counterexamples may be
omitted.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int},
  \ttp{Char}, or \ttp{Bool}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, Tritan Ravitch, the
anonymous reviewers of ICFP'13 (from which an earlier draft of this paper was
rejected), and especially John Hughes.  Their deep insights and bug catches
dramatically improved the presentation and content.
